Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
After spellcheck

Anyone who analyzes person-data may run into a missing data problem.
Missing data is not only ubiquitous, but also tedious. If a dataset
contains just one incomplete observation, statistical inferences are
undefined and will not produce any results. To circumvent this, many
statistical packages employ list-wise deletion by default (i.e.,
ignoring incomplete observations). Unfortunately, this \emph{ad hoc}
solution may yield wildly invalid results \citep{buur18}. An alternative
is to `impute' (i.e., fill in) the missing values in the incomplete
observations. Subsequently, statistical inferences can be performed on
the completed dataset. By repeating this process several times, a
distribution of plausible results may be obtained, which reflects the
uncertainty in the data due to missingness. This technique is known as
`multiple imputation' \citep[MI;][]{rubin76}. MI yields unbiased and
confidence valid estimates of the true---but missing---data inference
under many circumstances \citep{buur18}.

@@

In this paper we investigate how non-convergence in iterative imputation
algorithms may be diagnosed, how well these methods perform, and at
which point convergence may safely be assumed. For reasons of brevity,
we only focus on the MI algorithm implemented in the world-leading MI
software: the \texttt{R} \citep{R} package \texttt{mice} \citep{mice}.
The convergence properties of this MI algorithm are investigated through
model-based simulation. The results of this simulation study are
guidelines for assessing convergence of MI algorithms, which will aid
applied researchers in drawing valid inference from incomplete datasets.

@@

The two challenges for convergence of iterative algorithms are mixing
and stationarity \citep{gelm13}. Without mixing, imputation chains do
not intermingle nicely. Without stationarity, there is trending within
imputation chains, see Figure \ref{fig:non-conv}. \textbf{Explain what
we see, namely example by \citet{buur18} reproduced, showing the
traceplots of chain means for some variable. The first plot is typical
convergence of MICE, the second is pathological non-convergence because
of a mis-specified imputation model. Each line is an imputation. In the
first plot, the lines intermingle nicely and there is little trending.
In the second, there is a lot of trending and the lines do not cross
much. Importantly, the chain means at the last iteration (the imputed
value per \(m\)) are very different between the two plots. The algorithm
with the mis-specified model yields imputed values that are on average a
magnitude two larger than those of the typically converged algorithm.
This shows the importance of reaching convergend states in iterative
imputation algorithms.}

@@

Following the same notation, we define autocorrelation as the
correlation between two subsequent \(\theta\)-values within the same
chain \citep[p.~147]{lync07}. In this study we only consider \(AC\) at
lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\)
iteration of the same chain.

@@

We can interpret \(AC\)-values as a measure of stationarity. If
\(AC\)-values are close to zero, there is no dependence between
subsequent samples within imputation chains. Negative \(AC\)-values
indicate divergence within imputation chains. \textbf{Subsequent sampled
values within each imputation chain are less alike.} Positive
\(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent
iterations are similar, trending may occur. Negative \(AC\)-values show
no threat to the stationarity component of convergence. On the contrary
even---negative \(AC\)-values indicate that \(\theta\)-values of
subsequent iterations diverge from one-another, which may increase the
variance of \(\theta\) and speed up convergence. As convergence
diagnostic, the interest is therefore in positive \(AC\)-values.
\textbf{Maybe remove:} Moreover, the magnitude of \(AC\)-values may be
evaluated statistically, but that is outside of this note's scope.

@@

\textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
non-convergence diagnostics for MICE algorithms, the following condtion
must hold when applied on the two algorithms plotted above: the methods
indicate worse performance for the mis-specified model with pathological
non-convergence (i.e., higher \(\widehat{R}\)- and \(AC\)-values than
the typical performance). And additionally the methods should reflect
the increasing convergence in the typical convergence situation as the
number of iterations goes up.}

\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
the chain means are plotted again, now together. So are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up in the typical convergence situation. The superior
performance of the typical convergence over the pathological
non-convergnce is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function stats::acf(). When
we look at this panel, we must conclude that something weird is going
on. The \(AC\)-values indicate equal performance (upto \(t=5\)) for the
typical convergence and the pathological non-convergnce, while there is
obvious trending in the latter. Moreover, the best convergence (as
indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending upto iteration number seven. After consulting the documentation
on stats::acf(), we conclude that this implementation of \(AC\) is not
suitable for iterative imputation algorithms. There is a correction
factor for a mathematical shortcut that works in the limit. According to
\ldots{} the function works when the number of iterations \textgreater{}
50. The default number of iterations in mice, however, is 5. Therefore,
we compute \(AC\) manually, see panel D. The AC values in this plot do
meet the requirements (\ldots) and will therefore be used in the
simulation study.}

@@

Convergence is reached when there is no dependency between subsequent
iterations of imputation chains (\(AC = 0\)), and chains intermingle
such that the only difference between the chains is caused by the
randomness induced by the algorithm (\(\widehat{R} = 1\)). We expect
that a completely convergenced state will not be reached.

@@

The two challenges of convergence are stationarity and mixing. Mixing
and stationarity can be inspected visually, by evaluating traceplots of
scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
variances). As \citet{buur18} describes, users can also specify a
model-specific scalar summary (e.g., a regression coefficient). A
user-specified scalar summary, however, is not universal to all complete
data problems. Therefore, as inspired by \citep{mack03}, we will use
introduce a \(\theta\) that summarizes the multivariate state space of
the algorithm. Namely, the first eigenvalue of the variance-covariance
matrix of the \(M\) completed datasets.

@@

The performance of two methods (\(\widehat{R}\) and \(AC\)) on each of
these scalar summaries of interest will be evaluated with several
estimands (quantities of scientific interest; \(Q\)s). It is assumed
that when \(\bar{Q}\) (the pooled result across imputations) is an
unbiased (and confidence valid) estimate of \(Q\), the algorithm is
sufficiently converged. For each scalar \(\theta\), one estimand \(Q\)
is defined that may be of interest in empirical research. The
\(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s are
evaluated against the bias in univariate mean estimates (\textbf{all
variables or just for Y??}). Similarly, the performance measure for
\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
estimated standard deviations. To evaluate the performance of
\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
coefficient of determination, \(R^2\). Additionally, we will evaluate
the estimated regression coefficients, the coverage rate of the CI95\%
of the regression estimates, and the CI length (\textbf{see definitions
in old Methods}). \textbf{remove the term performance measure here and
add an extra alinea to define performance measures as bias in all
estimands, and coverage rate and CI length of regression coefficients.}

\textbf{Move this to Results:} As expected, conditions with a higher
proportion of missingness and/or a lower number of iterations show more
signs of non-convergence, as indicated by more extreme bias in the
estimated \(Q\)s. Roughly speaking this means that MICE is indeed not
converged as \(t=1\), and converges gradually as \(t\) increases. The
point at wich an additional iteration does not lead to an improvement of
the estimates depends on the difficulty of the missingness
problem---between XYZ and XYZ. This means that the algorithm has
converged sufficiently (\textbf{under the current specifications}).

@@

Evaluation with the performance measures shows that are had an
autocorrelation are conservative: they indicate signs of non-convergence
in conditions where \(\bar{Q}\)s are unbiased and or confidence valid
estimates of \(Q\).

@@

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), but in MCMC algorithms
these are very unlikely (because it's a distribution, not a point to
converge to). In practice thresholds are used to diagnose
non-convergence. Usually, non-convergence is diagnosed when
\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
performed to assess whether \(AC\) is significantly different from zero.
Performance measures are the same as above: unbiased, confidence valid
estimates.

\textbf{Move this to Results:} As expected, complete convergence
(\(\widehat{R}=1\) and \(AC=0\)) is not observed. But what about the
thresholds for practice? \(\widehat{R}\) \textless1.2 is not stringent
enough, because conditions in which \(\widehat{R}\) is smaller than 1.2
do not all yield unbiased estimates of the \(Q\)s. Moreover, there is a
dip in the \(\widehat{R}\) values, after which \(\widehat{R}\)
\textgreater{} 1.2 again. If the algorithms would be terminated at the
iteration where \(\widehat{R}\) \textless{} 1.2 occurs for the first
time, the increase in \(\widehat{R}\) values might be missed. The
threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative with
respect to the performance measures. But it seems OK. \(\widehat{R}\)
\textless{} 1.01 is too stringent compared to the performance measures.
And it is not obtained for any number of iterations (\textbf{but may be
necessary in more complex complete data models???}).

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence by: 1) terminating the MICE algorithm at different
imputation chain lengths (\(t = 1, 2, ..., 100\)), and 2) varying the
missingness proportions (\(p_{miss} = .05, .25, .50, .75, .95\)). The
assumption underlying the different number of iterations is that the
algorithm generally does reach convergence at \(t=1\), because the
initial values are sampled randomly from the set of observed datapoints.
\textbf{As the number of iterations goes up, the imputation chains will
become independent of the initial values until the point at which an
extra added iteration does not lead to a more converged state.} The
second set of experimental conditions under consideration--the
missingness proportions--are chosen to reflect the difficulty of the
missingness problem. The inherent assumption is that low missingness
proportions will lead to quick algorithmic convergence, since there is a
lot of information in the observed data. Higher missingness proportions
then cause slower convergence. However, if the fraction of missing
information is very high, there is so little information in the data
that the random component in the algorithm will take the overhand and a
stable but very uncertain (high variance) point will be reached.

@@

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim  \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} For the purpose of this study, sampling variance is not of
interest. Therefore, a single complete set may serve as comparative
truth in all simulation runs \citep{vinknd}. A finite population of
\(N=1000\) is simulated using the function \texttt{mvtnorm::rmvnorm()}.
Subsequently, a fourth variable is constructed to serve as dependent
variable in a multiple linear regression problem. Let \[
Y_i =  1 + 2X_{1i} + .5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2, ..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

We consider four quantities of scientific interest \citep[`conceptual
estimands';][]{morr19}, namely \ldots{} solve a multiple linear
regression problem, where dependent variable \(Y\) is regressed on
independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
notation with betas here, and add what the quantity/-ies of scientific
interest is/are!}):

@@

The complete data is `amputed' once for each simulation repetition with
function \texttt{mice::ampute()}. (\textbf{change this to reflect
current simulation set-up with 5, 25, 50, 75, and 95\% of cases having
missing data:} The missingness is univariate, and the probability to be
missing is the same for all four variables, namely 20\%
(\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This leaves 20\% of the rows
completely observed). This study only considers only an MCAR missingness
mechanism. Therefore, results may not be extrapolated to other missing
data problems. Proper performance of the convergence diagnostics under
MCAR is necessary but not sufficient to demonstrate appropriateness of
\(\widehat{R}\) and \(AC\) as convergence diagnostics. This is just a
proof of concept.

Missing datapoints are imputed with the function \texttt{mice::mice()}.
All MI procedures are performed with Bayesian linear regression
imputation (\texttt{method\ =\ "norm"}), and five imputation chains
(\texttt{m\ =\ 5}). The number of iterations varies between simulation
conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).

\(\widehat{R}\) (\textbf{of what?? i.e., chain means and chain variances
of each variable}) is computed by implementing Vehtari et al.'s
\citeyearpar{veht19} recommendations. \(AC\) (\textbf{of what? and how
is \(AC\) aggregated across imputations? i.e., mean in latest sim}) is
computed with function \texttt{stats::acf()}.

To estimate the quantity of scientific interest, \(Q\), we perform
multiple linear regression on each completed dataset with the function
\texttt{stats::lm()}. We obtain an estimated regression coefficient per
imputation, which are pooled into a single estimate, \(\bar{Q}\). We use
the function \texttt{mice::pool()} to estimate u bar (\textbf{check
definition!}) according to Rubin's \citeyearpar{rubin87} rules, and
subsequently implement finite population pooling conform \citet{vink14}.

We compute bias as the difference between \(Q\) and \(\bar{Q}\).
Confidence interval width (CIW) is defined as the difference between the
lower and upper bound of the 95\% confidence interval (CI95\%) around
\(\bar{Q}\). We compute the CI95\% bounds as

\[\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}},\]

where \(t_{(m-1)}\) is the quantile of a \(t\)-distribution with \(m-1\)
degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
pooled variance estimate. \textbf{Why track CIW?} Under-estimating the
variance of \(\bar{Q}\) may yield spurious inferences. From bias and
CIW, we calculate empirical coverage rates. Coverage rate is the
proportion of simulations in which \(Q\) is between the bounds of the
CI95\% around \(\bar{Q}\).

@@

@@

The bias in the estimates of the variable means show little to no
difference between simulation conditions. It doesn't seem to matter how
many iterations you use in the mice algorithm, the estimates are
unbiased. Similarly, the bias in the estimated variances is more or less
stable across simulation conditions. These univariate quantities appear
to be unaffected by the number of iterations.

When applied to the imputation chain means, \(\widehat{R}\) indicates
that the mice algorithm does not reach a converged state
(\(\widehat{R}\) = 1) in any of the simulation conditions. Neither is
the most recent recommended threshold reached (\(\widehat{R}\)
\textless{} 1.01). The conventional \(\widehat{R}\) threshold of 1.2 is
reached in simulation conditions \(T = 3\) and \(T > 6\). (With the
default number of iterations (maxit = 5), this dip in \(\widehat{R}\)
values would be spotted, so it is no problem.) The point at which an
extra iteration does not seem to improve the \(\widehat{R}\) value is
around \(T=30\).

Auto-correlations indicate no sign of trending within imputation chains.
In most simulation conditions \(AC\) is smaller than or about equal to
zero. Across simulation conditions, however, the autocorrelation curve
does not trend towards 0. Autocorrelation values plateau off at a value
of around .1 . This is a small positive autocorrelation, which would
indicate some trending within chains.

The \(\widehat{R}\) and \(AC\) values for the imputation chain variances
show equal trends and are therefore not discussed separately. Taken
together, univariate estimates seem robust across simulation conditions.
There is no clear effect of the number of iterations on the bias in
these estimates, while the convergence diagnostics indicate that the
algorithm did not reach a completely converged state (yet).

@@

There is a clear bias in the regression estimates in simulation
conditions where the number of iterations is smaller than four. In
simulation conditions where \(T > 5\) there is little to no bias in the
estimated regression coefficients. In most simulation conditions nominal
coverage is obtained (i.e., coverage rates of .95) for the confidence
intervals around the regression coefficients. Conditions with only one
or two iterations show some under-coverage. Since confidence interval
width is stable across conditions, the under-coverage may be attributed
to the bias in the estimated regression coefficients.

If we look at the estimated proportion of explained variance in outcome
variable Y we see that the coefficient of determination is underestimate
estimated in conditions where the number of iterations is equal to two
or less, and slightly overestimated in conditions where the number of
iterations is equal to three or more.

@@

--\textgreater{}

@@

\textbf{Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across
imputations, not iterations (unique for imp data??)}

\textbf{If I don't include it in my study, also: Significance of \(AC\)
values}

>>>>

Anyone who analyzes person-data may run into a missing data problem.
Missing data is not only ubiquitous but also tedious. If a dataset
contains just one incomplete observation, statistical inferences are
undefined and will not produce any results. To circumvent this, many
statistical packages employ list-wise deletion by default (i.e.,
ignoring incomplete observations). Unfortunately, this \emph{ad hoc}
solution may yield wildly invalid results \citep{buur18}. An alternative
is to `impute' (i.e., fill in) the missing values in the incomplete
observations. Subsequently, statistical inferences can be performed on
the completed dataset. By repeating this process several times, a
distribution of plausible results may be obtained, which reflects the
uncertainty in the data due to missingness. This technique is known as
`multiple imputation' \citep[MI;][]{rubin76}. MI yields unbiased and
confidence valid estimates of the true---but missing---data inference
under many circumstances \citep{buur18}.

@@

In this paper, we investigate how non-convergence in iterative
imputation algorithms may be diagnosed, how well these methods perform,
and at which point convergence may safely be assumed. For reasons of
brevity, we only focus on the MI algorithm implemented in the
world-leading MI software: the \texttt{R} \citep{R} package
\texttt{mice} \citep{mice}. The convergence properties of this MI
algorithm are investigated through model-based simulation. The results
of this simulation study are guidelines for assessing convergence of MI
algorithms, which will aid applied researchers in drawing valid
inference from incomplete datasets.

@@

The two challenges for convergence of iterative algorithms are mixing
and stationarity \citep{gelm13}. Without mixing, imputation chains do
not intermingle nicely. Without stationarity, there is trending within
imputation chains, see Figure \ref{fig:non-conv}. \textbf{Explain what
we see, namely example by \citet{buur18} reproduced, showing the
traceplots of chain means for some variable. The first plot is typical
convergence of MICE, the second is pathological non-convergence because
of a mis-specified imputation model. Each line is an imputation. In the
first plot, the lines intermingle nicely and there is little trending.
In the second, there is a lot of trending and the lines do not cross
much. Importantly, the chain means at the last iteration (the imputed
value per \(m\)) are very different between the two plots. The algorithm
with the mis-specified model yields imputed values that are on average a
magnitude two larger than those of the typically converged algorithm.
This shows the importance of reaching converged states in iterative
imputation algorithms.}

@@

Following the same notation, we define autocorrelation as the
correlation between two subsequent \(\theta\)-values within the same
chain \citep[p.~147]{lync07}. In this study, we only consider \(AC\) at
lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\)
iteration of the same chain.

@@

We can interpret \(AC\)-values as a measure of stationarity. If
\(AC\)-values are close to zero, there is no dependence between
subsequent samples within imputation chains. Negative \(AC\)-values
indicate divergence within imputation chains. \textbf{Subsequent sampled
values within each imputation chain are less alike.} Positive
\(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent
iterations are similar, trending may occur. Negative \(AC\)-values show
no threat to the stationarity component of convergence. On the contrary
even---negative \(AC\)-values indicate that \(\theta\)-values of
subsequent iterations diverge from one another, which may increase the
variance of \(\theta\) and speed up convergence. As convergence
diagnostic, the interest is therefore in positive \(AC\)-values.
\textbf{Maybe remove:} Moreover, the magnitude of \(AC\)-values may be
evaluated statistically, but that is outside of this note's scope.

@@

\textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
non-convergence diagnostics for MICE algorithms, the following condition
must hold when applied on the two algorithms plotted above: the methods
indicate worse performance for the mis-specified model with pathological
non-convergence (i.e., higher \(\widehat{R}\)- and \(AC\)-values than
the typical performance). And additionally the methods should reflect
the increasing convergence in the typical convergence situation as the
number of iterations goes up.}

\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
the chain means are plotted again, now together. So are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up in the typical convergence situation. The superior
performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function stats::acf(). When
we look at this panel, we must conclude that something weird is going
on. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for the
typical convergence and the pathological non-convergence, while there is
obvious trending in the latter. Moreover, the best convergence (as
indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf(), we conclude that this implementation of
\(AC\) is not suitable for iterative imputation algorithms. There is a
correction factor for a mathematical shortcut that works in the limit.
According to\ldots{} the function works when the number of iterations
\textgreater{} 50. The default number of iterations in mice, however, is
5. Therefore, we compute \(AC\) manually, see panel D. The AC values in
this plot do meet the requirements (\ldots) and will therefore be used
in the simulation study.}

@@

Convergence is reached when there is no dependency between subsequent
iterations of imputation chains (\(AC = 0\)), and chains intermingle
such that the only difference between the chains is caused by the
randomness induced by the algorithm (\(\widehat{R} = 1\)). We expect
that a completely converged state will not be reached.

@@

The two challenges of convergence are stationarity and mixing. Mixing
and stationarity can be inspected visually, by evaluating traceplots of
scalar summaries of interest (\(\theta\)s; e.g., chain means and chain
variances). As \citet{buur18} describes, users can also specify a
model-specific scalar summary (e.g., a regression coefficient). A
user-specified scalar summary, however, is not universal to all complete
data problems. Therefore, as inspired by \citep{mack03}, we will propose
a \(\theta\) that summarizes the multivariate state space of the
algorithm. Namely, the first eigenvalue of the variance-covariance
matrix of the \(M\) completed datasets.

@@

The performance of two methods (\(\widehat{R}\) and \(AC\)) on each of
these scalar summaries of interest will be evaluated with several
estimands (quantities of scientific interest; \(Q\)s). It is assumed
that when \(\bar{Q}\) (the pooled result across imputations) is an
unbiased (and confidence valid) estimate of \(Q\), the algorithm is
sufficiently converged. For each scalar \(\theta\), one estimand \(Q\)
is defined that may be of interest in empirical research. The
\(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s are
evaluated against the bias in univariate mean estimates (\textbf{all
variables or just for Y??}). Similarly, the performance measure for
\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
estimated standard deviations. To evaluate the performance of
\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
coefficient of determination, \(R^2\). Additionally, we will evaluate
the estimated regression coefficients, the coverage rate of the CI95\%
of the regression estimates, and the CI length (\textbf{see definitions
in old Methods}). \textbf{remove the term performance measure here and
add an extra paragraph to define performance measures as bias in all
estimands, and coverage rate and CI length of regression coefficients.}

\textbf{Move this to Results:} As expected, conditions with a higher
proportion of missingness and/or a lower number of iterations show more
signs of non-convergence, as indicated by more extreme bias in the
estimated \(Q\)s. Roughly speaking this means that MICE is indeed not
converged as \(t=1\), and converges gradually as \(t\) increases. The
point at which an additional iteration does not lead to an improvement
of the estimates depends on the difficulty of the missingness
problem---between XYZ and XYZ. This means that the algorithm has
converged sufficiently (\textbf{under the current specifications}).

@@

Evaluation with the performance measures shows that \(\widehat{R}\) and
autocorrelation are conservative: they indicate signs of non-convergence
in conditions where \(\bar{Q}\)s are unbiased and or confidence valid
estimates of \(Q\).

@@

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), but in MCMC algorithms
these are very unlikely (because it's a distribution, not a point to
converge to). In practice, thresholds are used to diagnose
non-convergence. Usually, non-convergence is diagnosed when
\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
performed to assess whether \(AC\) is significantly different from zero.
Performance measures are the same as above: unbiased, confidence valid
estimates.

\textbf{Move this to Results:} As expected, complete convergence
(\(\widehat{R}=1\) and \(AC=0\)) is not observed. But what about the
thresholds for practice? \(\widehat{R}\) \textless1.2 is not stringent
enough, because conditions in which \(\widehat{R}\) is smaller than 1.2
do not all yield unbiased estimates of the \(Q\)s. Moreover, there is a
dip in the \(\widehat{R}\) values, after which \(\widehat{R}\)
\textgreater{} 1.2 again. If the algorithms would be terminated at the
iteration where \(\widehat{R}\) \textless{} 1.2 occurs for the first
time, the increase in \(\widehat{R}\) values might be missed. The
threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative in
comparison with the performance measures. But it seems OK.
\(\widehat{R}\) \textless{} 1.01 is too stringent compared to the
performance measures. And it is not obtained for any number of
iterations (\textbf{but may be necessary in more complex complete data
models???}).

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence by: 1) terminating the MICE algorithm at different
imputation chain lengths (\(t = 1, 2,..., 100\)), and 2) varying the
missingness proportions (\(p_{miss} =.05,.25,.50,.75,.95\)). The
assumption underlying the different number of iterations is that the
algorithm generally does reach convergence at \(t=1\), because the
initial values are sampled randomly from the set of observed datapoints.
\textbf{As the number of iterations goes up, the imputation chains will
become independent of the initial values until the point at which an
extra added iteration does not lead to a more converged state.} The
second set of experimental conditions under consideration--the
missingness proportions--are chosen to reflect the difficulty of the
missingness problem. The inherent assumption is that low missingness
proportions will lead to quick algorithmic convergence, since there is a
lot of information in the observed data. Higher missingness proportions
then cause slower convergence. However, if the fraction of missing
information is very high, there is so little information in the data
that the random component in the algorithm will take the overhand and a
stable but very uncertain (high variance) point will be reached.

@@

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} In this study, sampling variance is not of interest.
Therefore, a single complete set may serve as comparative truth in all
simulation runs \citep{vinknd}. A finite population of \(N=1000\) is
simulated using the function \texttt{mvtnorm::rmvnorm()}. Subsequently,
a fourth variable is constructed to serve as dependent variable in a
multiple linear regression problem. Let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

We consider four quantities of scientific interest \citep[`conceptual
estimands';][]{morr19}, namely\ldots{} solve a multiple linear
regression problem, where dependent variable \(Y\) is regressed on
independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
notation with betas here, and add what the quantity/-ies of scientific
interest is/are! Namely, the descriptive statistics of all variables
(mean and standard deviation), \(\beta_2\), and the percentage of
variance in \(Y\) explained (coefficient of determination; \(R^2\))}):

@@

The complete data is `amputed' once for each simulation repetition with
function \texttt{mice::ampute()}. (\textbf{change this to reflect
current simulation set-up with 5, 25, 50, 75, and 95\% of cases having
missing data:} The missingness is univariate, and the probability to be
missing is the same for all four variables, namely 20\%
(\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This leaves 20\% of the rows
completely observed). This study only considers only an MCAR missingness
mechanism. Therefore, results may not be extrapolated to other missing
data problems. Proper performance of the convergence diagnostics under
MCAR is necessary but not sufficient to demonstrate the appropriateness
of \(\widehat{R}\) and \(AC\) as convergence diagnostics. Convergence
diagnostics should therefore at least apply to the MCAR situation,
before more complex missingness should be explored.

Missing datapoints in \(y\) are imputed with the function
\texttt{mice::mice()}. All MI procedures are performed with Bayesian
linear regression imputation (\texttt{method\ =\ "norm"}), and five
imputation chains (\texttt{m\ =\ 5}). The number of iterations varies
between simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).
Each repetition of the simulation thus starts from 1 complete dataset,
then splits into 5 missingness conditions (\(p =.05,.25,.5,.75,.95\)),
each of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\))
for each iteration condition (\(t = 1, 2,..., 100\)). From these
\(5\times5\times100\) imputations, we extract the \(\theta\)s to apply
\(\widehat{R}\) and \(AC\) on (\textbf{i.e., chain means, chain
variances, \(\beta\)s per imputation, and the first eigenvalue of the
variance-covariance matrix per imputation}). We calculate
\(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19}
recommendations, and \(AC\) as the correlation between the \(t^{th}\)
and the \((t+1)^{th}\) iteration. For each \(\theta\) we estimate the
corresponding scientific estimand \(Q\).

The estimator for each \(Q\) is \(\bar{Q}\)---the pooled aggregate of
the (\(\hat{Q}\))s across imputations. To calculate the \(\bar{Q}\)s, we
combine the observed data (\(y_{obs}\)) and the imputed data for each
imputation (\(y_{imp,m}\)) with the function \texttt{mice::complete()}.
Descriptive statistics are computed as the average across imputations
for each variable (\(\mu_j\) and \(\sigma_j\), where
\(j = Y, X_1, X_2, X_3\)). Estimated regression coefficients are
obtained with the function \texttt{stats::lm()} for each imputation, and
then pooled conform \citet{vink14}. The coefficient of determination is
estimated for each imputation, and pooled using
\texttt{mice::pool.r.squared()}.

The performance of \(\widehat{R}\) and \(AC\) is assessed by comparing
\(\bar{Q}\)s with \(Q\)s. For each \(Q\), we compute bias as
\(\bar{Q} - Q\). For \(Q=\beta_j\), we also compute the empirical
coverage rate (CR). CR is defined as the percentage of simulation
repetitions in which the 95\% confidence interval (CI) around
\(\bar{Q}\) covers the true estimand \(Q\). Let

\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\]

where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\)
degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
pooled variance estimate. \textbf{Remove this? } Under-estimating the
variance of \(\bar{Q}\) may yield spurious inferences. Confidence
interval width (CIW) is defined as the difference between the lower and
upper bound of the 95\% confidence interval (CI) around \(\bar{Q}\) From
bias and CIW, we calculate empirical coverage rates. Coverage rate is
the proportion of simulations in which \(Q\) is between the bounds of
the CI95\% around \(\bar{Q}\).

@@

For reasons of brevity, we only discuss the convergence diagnostics for
the \(Q\) with the worst performance in terms of bias. For the
descriptve statistics, the magnitude of the bias was the largest in
\(Y\) (i.e, \(j = Y\) in \(\mu_j\) and \(\sigma_j\)). For the
\(Q=\beta_j\), \(j=2\) showed the worst perfomance (i.e, the effect of
\(X_1\) on \(Y\).

@@

The bias in the estimates of the variable means shows little to no
difference between simulation conditions. It doesn't seem to matter how
many iterations you use in the mice algorithm, the estimates are
unbiased. Similarly, the bias in the estimated variances is more or less
stable across simulation conditions. These univariate quantities appear
to be unaffected by the number of iterations.

When applied to the imputation chain means, \(\widehat{R}\) indicates
that the mice algorithm does not reach a converged state
(\(\widehat{R}\) = 1) in any of the simulation conditions. Neither is
the most recent recommended threshold reached (\(\widehat{R}\)
\textless{} 1.01). \textbf{This is only for 75\% missingness:} The
conventional \(\widehat{R}\) threshold of 1.2 is reached in simulation
conditions \(T = 3\) and \(T > 6\). (With the default number of
iterations (maxit = 5), this dip in \(\widehat{R}\) values would be
spotted, so it is no problem.) The point at which an extra iteration
does not seem to improve the \(\widehat{R}\) value is around \(T=30\).

Autocorrelations indicate no sign of trending \textbf{(NOT correct
anymore, fluke of the stats::acf() function!!)} within imputation
chains. In most simulation conditions \(AC\) is smaller than or about
equal to zero. Across simulation conditions, however, the
autocorrelation curve does not trend towards 0. Autocorrelation values
plateau off at a value of around.1. This is a small positive
autocorrelation, which would indicate some trending within chains.

The \(\widehat{R}\) and \(AC\) values for the imputation chain variances
show equal trends to the chain means and are therefore not discussed
separately. Taken together, univariate estimates seem robust across
simulation conditions. There is no clear effect of the number of
iterations on the bias in these estimates, while the convergence
diagnostics indicate that the algorithm did not reach a completely
converged state (yet).

@@

\textbf{This is only true for 75\% missingness:} There is a clear bias
in the regression estimates in simulation conditions where the number of
iterations is smaller than four. In simulation conditions where
\(T > 5\) there is little to no bias in the estimated regression
coefficients. In most simulation conditions nominal coverage is obtained
(i.e., coverage rates of.95) for the confidence intervals around the
regression coefficients. Conditions with only one or two iterations show
some under-coverage. Since confidence interval width is stable across
conditions, the under-coverage may be attributed to the bias in the
estimated regression coefficients.

\textbf{This is only true for 75\% missingness:} If we look at the
estimated proportion of explained variance in outcome variable Y we see
that the coefficient of determination is underestimate estimated in
conditions where the number of iterations is equal to two or less, and
slightly overestimated in conditions where the number of iterations is
equal to three or more.

@@

@@

\textbf{Pmm, M(N)AR, Empirical data, Vignette, ShinyMICE, AC across
imputations, not iterations (new convergence diagnostic unique for
iterative imputation??).}

\textbf{If I don't include it in my study, also: Significance of \(AC\)
values.}
